Experience Report:

### 2010-03-17 doc/icfp-2010/main.tex
master -> compute ; handle -> master ; f -> worker / shutdown workers at exit

The main idea is to start with a generic function, with the
following interface: 
% TODO: rename this function?
\begin{ocaml}
  val compute : 
    f:('a -> 'b) -> 
    handle:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
% TODO be even more explicit about the introduction of the word `task'
This function takes two functions, \ocaml{f} and \ocaml{handle}, and a
list of \emph{tasks} of type \ocaml{('a * 'c) list}.
The function \ocaml{f} is applied to the first component for each task.
Whenever it completes, the task and the result are passed to the
function \ocaml{handle}. This may possibly generate a new list of
tasks. The process ends when all computations are done.

Function \ocaml{f} should be pure and is executed in parallel.
Function \ocaml{handle}, on the contrary, can be impure and is never
executed in parallel with other \ocaml{handle}s. Typically,
\ocaml{handle} stores and processes information locally.
The next section will describe the usage of this generic interface to
perform traditional map and fold operations.

@@

\item \textbf{master and workers are different programs, compiled with
    the same version of \Ocaml:} 
  in this case, we can no longer marshal closures but we can still
  marshal polymorphic values. As a consequence, 
  the \ocaml{compute} function is split into two
  polymorphic functions, to implement the master and workers respectively:
  \begin{ocaml}
val master : 
  handle:('a * 'c -> 'b -> ('a * 'c) list) -> 
  ('a * 'c) list -> unit
val worker : ('a -> 'b) -> unit -> unit
  \end{ocaml}

\item \textbf{master and workers are different programs, not even
    compiled with the same version of \Ocaml:} in this case, we can no
  longer use marshaling, so the
  \ocaml{compute} function is split into two monomorphic functions
  over strings: 
  \begin{ocaml}
val master : 
  handle:(string * 'c -> string -> (string * 'c) list) -> 
  (string * 'c) list -> unit
val worker : (string -> string) -> unit -> unit
  \end{ocaml}
  Note that the second component of each task is still polymorphic (of
  type \ocaml{'c} here).
\end{enumerate}

@@

These five functions are actually derived from the generic API.
The generality is achieved by implementing theses derivations as a functor.
This functor, \of{Derive}, is parameterized by the \of{compute}
function, as follows:
\begin{ocaml}
module Derive
  (X : sig
     val compute : 
       f:('a -> 'b) -> 
       handle:('a * 'c -> 'b -> ('a * 'c) list) ->
       ('a * 'c) list -> unit
   end) = struct ... end
\end{ocaml}
We now explain how each function from Figure~\ref{fig:derived} is
implemented using \of{compute} inside the functor body.

\paragraph{\of{map}.} 
Function \of{map} is easily implemented by associating an integer
index with each element (of type \of{'a}) of the input list. Thus the
list of tasks passed to \of{compute} is a list of pairs of type \of{'a * int}.
As soon as an intermediate result is available, it is passed to
\of{handle}, which stores it in a local table using the index as a key
and generates no new task. Whenever \of{compute} returns, we easily
recover the output list from the values stored in the table.

\paragraph{\of{map_local_fold}.} 
Function \of{map_local_fold} is implemented using a local reference
storing the current accumulator (of type \of{'c}).
As soon as an intermediate result is available, \of{handle} 
combines it with the accumulator using function \of{fold}. Whenever \of{compute}
returns, we simply return the current accumulator.

\paragraph{\of{map_remote_fold}.}
In this case, combining the accumulator with an intermediate result is
itself a task to be performed in parallel with \of{map} operations.
Thus, when an intermediate \of{map} result is available, the
accumulator may be already in use in a \of{fold} operation. Hence we
need to store the intermediate \of{map} result in a local table, to be
used as soon as the accumulator becomes available.
The \of{handle} function distinguishes between \of{map} and \of{fold}
results and, accordingly, updates the table and the accumulator. For a
\of{map} result, either we combine it with the accumulator in a new
task or we simply store it in the local table. For a \of{fold} result,
either we simply store it into the accumulator or we immediately
combine it with a pending \of{map} result from the table.
Whenever \of{compute} returns, the accumulator must be available and
we simply return it.

\paragraph{\of{map_fold_ac}.}
Function \of{map_fold_ac} is easier than the previous one, since we do
not need to store more than one intermediate result. Indeed, as soon
as two intermediate results are available, we immediately combine them
using a \of{fold} operation. Thus the \of{handle} function 
either stores it or combines it, depending on the availability/non
availability of the accumulator.
Whenever \of{compute} returns, the accumulator must be available and
we simply return it.

@@

The \of{Cores} module is implemented with \unix\ processes, using the
\of{fork} and \of{wait} system calls provided by the \of{Unix} library
of \Ocaml. The idea is pretty simple. 
The \of{compute} function maintains a global table of pending tasks
and keeps track of the number of idle cores.
Whenever there a pending task and an idle core, 
a new sub-process is created using \of{Unix.fork}; 
once completed, the sub-process marshals the result 
into a local file. The main process maintains a table
mapping each sub-process ID to the input task and the local file
name. It waits for any completed sub-process using \of{Unix.wait}
and recovers the result from the local file. Then function \of{handle}
is applied, which may generate new tasks.
The main loop can be depicted in the following way:
\begin{flushleft}
  \quad  \textbf{while} pending tasks $\lor$ pending sub-processes \\
  \quad  \quad \textbf{while} pending tasks $\land$ idle cores \\
  \quad  \quad \quad create new sub-process for some task \\
  \quad  \quad \textbf{wait} for any completed sub-process \\
  \quad  \quad \quad push new tasks generated by handle \\
\end{flushleft}

@@

As explained in Section~\ref{sec:API}, the \of{Network} module is
actually providing three different implementations, according to three
different execution scenarios (namely, \of{Same}, \of{Poly} and
\of{Mono}). There is actually a single, generic implementation, which
is then instantiated in three different ways.
The generic implementation assumes that we are only transmitting
strings over the network, as evident from \of{Assign} and
\of{Completed} messages in the protocol section~\ref{sec:protocol}.
Therefore, the signature for the generic implementation is as follows:
\begin{ocaml}
  generic_compute:
    assign_job:('a -> string * string) ->
    handle:('a * 'c -> string -> ('a * 'c) list)  ->
    ('a * 'c) list -> unit
\end{ocaml}
The first argument \of{assign_job} is a function indicating 
the two strings to pass inside the \of{Assign} message of the protocol.
We can now distinguish the three cases:
\begin{description}
\item[\of{Same}:] in this case, \of{assign_job} is simply returning
  the marshaled closure together with the marshaled value.
\item[\of{Poly}:] in this case, \of{assign_job} is simply
  returning the marshaled value (the first string being meaningless in
  this context).
\item[\of{Mono}:] in this case, \of{assign_job} is immediately
  returning its argument (the first string being meaningless in
  this context).
\end{description}

@@

% LocalWords:  parallelize functor parameterized indices endianness

>>>>

The main idea is to start with a generic function, with the
following interface: 
% TODO: rename this function?
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
% TODO be even more explicit about the introduction of the word `task'
This function takes two functions, \ocaml{worker} and \ocaml{master}, and a
list of \emph{tasks} of type \ocaml{('a * 'c) list}.
The function \ocaml{worker} is applied to the first component for each task.
Whenever it completes, the task and the result are passed to the
function \ocaml{master}. This may possibly generate a new list of
tasks. The process ends when all computations are done.

Function \ocaml{worker} should be pure and is executed in parallel.
Function \ocaml{master}, on the contrary, can be impure and is never
executed in parallel with other \ocaml{master}s. Typically,
\ocaml{master} stores and processes information locally.
The next section will describe the usage of this generic interface to
perform traditional map and fold operations.

@@

\item \textbf{master and workers are different programs, compiled with
    the same version of \Ocaml:} 
  in this case, we can no longer marshal closures but we can still
  marshal polymorphic values. As a consequence, 
  the \ocaml{compute} function is split into two
  polymorphic functions, to implement the master and workers respectively:
  \begin{ocaml}
val master : 
  ('a * 'c -> 'b -> ('a * 'c) list) -> 
  ('a * 'c) list -> unit
val worker : ('a -> 'b) -> unit -> unit
  \end{ocaml}

\item \textbf{master and workers are different programs, not even
    compiled with the same version of \Ocaml:} in this case, we can no
  longer use marshaling, so the
  \ocaml{compute} function is split into two monomorphic functions
  over strings: 
  \begin{ocaml}
val master : 
  (string * 'c -> string -> (string * 'c) list) -> 
  (string * 'c) list -> unit
val worker : (string -> string) -> unit -> unit
  \end{ocaml}
  Note that the second component of each task is still polymorphic (of
  type \ocaml{'c} here).
\end{enumerate}

@@

These five functions are actually derived from the generic API.
The generality is achieved by implementing theses derivations as a functor.
This functor, \of{Derive}, is parameterized by the \of{compute}
function, as follows:
\begin{ocaml}
module Derive
  (X : sig
     val compute : 
       worker:('a -> 'b) -> 
       master:('a * 'c -> 'b -> ('a * 'c) list) ->
       ('a * 'c) list -> unit
   end) = struct ... end
\end{ocaml}
We now explain how each function from Figure~\ref{fig:derived} is
implemented using \of{compute} inside the functor body.

\paragraph{\of{map}.} 
Function \of{map} is easily implemented by associating an integer
index with each element (of type \of{'a}) of the input list. Thus the
list of tasks passed to \of{compute} is a list of pairs of type \of{'a * int}.
As soon as an intermediate result is available, it is passed to
\of{master}, which stores it in a local table using the index as a key
and generates no new task. Whenever \of{compute} returns, we easily
recover the output list from the values stored in the table.

\paragraph{\of{map_local_fold}.} 
Function \of{map_local_fold} is implemented using a local reference
storing the current accumulator (of type \of{'c}).
As soon as an intermediate result is available, \of{master} 
combines it with the accumulator using function \of{fold}. Whenever \of{compute}
returns, we simply return the current accumulator.

\paragraph{\of{map_remote_fold}.}
In this case, combining the accumulator with an intermediate result is
itself a task to be performed in parallel with \of{map} operations.
Thus, when an intermediate \of{map} result is available, the
accumulator may be already in use in a \of{fold} operation. Hence we
need to store the intermediate \of{map} result in a local table, to be
used as soon as the accumulator becomes available.
The \of{master} function distinguishes between \of{map} and \of{fold}
results and, accordingly, updates the table and the accumulator. For a
\of{map} result, either we combine it with the accumulator in a new
task or we simply store it in the local table. For a \of{fold} result,
either we simply store it into the accumulator or we immediately
combine it with a pending \of{map} result from the table.
Whenever \of{compute} returns, the accumulator must be available and
we simply return it.

\paragraph{\of{map_fold_ac}.}
Function \of{map_fold_ac} is easier than the previous one, since we do
not need to store more than one intermediate result. Indeed, as soon
as two intermediate results are available, we immediately combine them
using a \of{fold} operation. Thus the \of{master} function 
either stores it or combines it, depending on the availability/non
availability of the accumulator.
Whenever \of{compute} returns, the accumulator must be available and
we simply return it.

@@

The \of{Cores} module is implemented with \unix\ processes, using the
\of{fork} and \of{wait} system calls provided by the \of{Unix} library
of \Ocaml. The idea is pretty simple. 
The \of{compute} function maintains a global table of pending tasks
and keeps track of the number of idle cores.
Whenever there a pending task and an idle core, 
a new sub-process is created using \of{Unix.fork}; 
once completed, the sub-process marshals the result 
into a local file. The main process maintains a table
mapping each sub-process ID to the input task and the local file
name. It waits for any completed sub-process using \of{Unix.wait}
and recovers the result from the local file. Then function \of{master}
is applied, which may generate new tasks.
The main loop can be depicted in the following way:
\begin{flushleft}
  \quad  \textbf{while} pending tasks $\lor$ pending sub-processes \\
  \quad  \quad \textbf{while} pending tasks $\land$ idle cores \\
  \quad  \quad \quad create new sub-process for some task \\
  \quad  \quad \textbf{wait} for any completed sub-process \\
  \quad  \quad \quad push new tasks generated by \of{master} \\
\end{flushleft}

@@

As explained in Section~\ref{sec:API}, the \of{Network} module is
actually providing three different implementations, according to three
different execution scenarios (namely, \of{Same}, \of{Poly} and
\of{Mono}). There is actually a single, generic implementation, which
is then instantiated in three different ways.
The generic implementation assumes that we are only transmitting
strings over the network, as evident from \of{Assign} and
\of{Completed} messages in the protocol section~\ref{sec:protocol}.
Therefore, the signature for the generic implementation is as follows:
\begin{ocaml}
  generic_master:
    assign_job:('a -> string * string) ->
    master:('a * 'c -> string -> ('a * 'c) list)  ->
    ('a * 'c) list -> unit
\end{ocaml}
The first argument \of{assign_job} is a function indicating 
the two strings to pass inside the \of{Assign} message of the protocol.
We can now distinguish the three cases:
\begin{description}
\item[\of{Same}:] in this case, \of{assign_job} is simply returning
  the marshaled closure together with the marshaled value.
\item[\of{Poly}:] in this case, \of{assign_job} is simply
  returning the marshaled value (the first string being meaningless in
  this context).
\item[\of{Mono}:] in this case, \of{assign_job} is immediately
  returning its argument (the first string being meaningless in
  this context).
\end{description}

@@

% LocalWords:  parallelize functor parameterized indices endianness monomorphic
